Probabilistic Topic Modelling for Spatial Smoothing in Mass Spectrometry Imaging

### 2016-12-18 docs/MSci_proposal/mprop.tex
the final review up to the evaluation.

% other potentially useful packages
\usepackage{amssymb,amsmath}
%\usepackage{url}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Probabilistic Topic Modelling for Spatial Smoothing in Mass Spectrometry Imaging}
\author{Arijus Pleska}
\date{Date of submission placed here}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\par In this project, the application domain of MSI is set to be \textit{metabolomics} -- it is the scientific study of chemical processes which involve metabolites (small molecules) and metabolomes (sets of small molecules). In the settings of metabolomics, we utilise \text{mass spectrometry} (MS) to obtain the distribution of chemical entities within a sample. By utilising MSI, we can visualise the distributions of the ions (or ion sets) of different masses. For instance, Figure \ref{fig:b_and_e} below represents a captured sample as an image, where each picture's pixel is a small region within the sample. 
\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{b_and_e}
  \caption{MSI application on a sample with the letters b and e.}
  \label{fig:b_and_e}
\end{figure}
Note that each pixel contains a distribution of chemical entities as well. In Figure \ref{fig:b_and_e} we have chosen the range of mass to reflect the ink used to inscribe the letters. That is, in each pixel we have summed the intensities of the chemical entities meeting the mass criteria. Then, each pixel has been assigned an intensity value. Therefore, the sample could be expressed as the latter image. An open problem in metabolomics is to find the sets of chemical entities  inducing valid visual patterns. Ultimately, the discovered biological patterns could be used to suggest and measure the likelihood of a particular chemical process occurrence. For a rigorous discussion on machine learning applications in metabolomics, the reader can consult the recent survey conducted by Alonso et al. \cite{alonso-et-al}.

@@

% Unsupervisder learning
% - Topic moels
% - Probabilistic models
% - LDA:
% -- General
% -- Metabolomics
\par Since this proposal is focused on utilising spatial smoothing to reduce the impact of noise in MSI data sets, we will familiarise with the machine learning techniques and problems tackled in the MSI domain. The objective to discover novel patterns can be treated as a problem of \textit{unsupervised} machine learning. That is, we have no initial settings in how the patterns are expected to look like. However, upon appropriate optimisation, the unsupervised machine learning techniques are capable to learn the patterns on their own. To be more specific, we will utilise the techniques of \textit{topic modelling} (in this case, a pattern is referred by a topic). Also, we are focusing on the \textit{probabilistic} (statistical) models, more specifically, \textit{latent Dirichlet allocation} (LDA). First of all, the successor LDA-like models are treated as a state-of-the-art method in tackling semantic-analysis type problems. For more detailed review on the LDA applications, the reader can consult the survey conducted by Blei \cite{blei_2012}. Also, Hooft et al. \cite{hooft} has recently addressed the applicability of LDA in MSI data sets. This proposal is directly influenced by the latter study: we will utilise the LDA-like models to reduce the impact of noise in the MSI data sets.

@@

% Objectives
% - Research in the state-of-the-art methodology.
% - Reducing the impact of noise in metabolomics.
% - Providing a general-purpose machine learning model.
\par The primary goal of this research is to expand the consensus on whether topic modelling is a valid approach to apply spatial smoothing in MSI data sets. To be more specific, we set our objectives to the following:
\begin{enumerate}
    \item Optimise the general topic modelling methods for MSI data sets;
    \item Successfully apply spatial smoothing in MSI data sets;
    \item Give a basis to a general topic modelling method for error correction.
\end{enumerate}
Briefly, for the first objective, we will conduct a survey on LDA-like topic modelling methodology. This survey will distinguish the potential models for the applications in MSI. The second objective focuses on optimising these potential models for the particular task of spatial smoothing in MSI data sets. Finally, the third objective can be treated as a general contribution to the field of machine learning, since the discovered spatial smoothing techniques might be applicable in other domains.

% Hypothesis
% - Topic modelling can improve the quality of metabolomics data sets.
% - The negation of the hypothesis would contribute in terms of reassessing the complexity of the problem.
\par The main project hypothesis -- the LDA-like models can detect and correct the degeneracies of MSI data sets -- can be interpreted as the spatial smoothing of MSI data sets. Both cases of proving and disproving this hypothesis would contribute to the research in MSI. The successful outcome would lead to the potential applications discussed in the following paragraph, whereas the unsuccessful outcome would set up additional requirements to tackle the issue of noisiness in MSI data sets.

% Results: tools to annotate the metabolites > basis for generative error correction 
% - Expanding the application of existing machine learning models.
% - Advancing the current state-of-the-art in metabolomics.
% - Contributing to machine learning in general.
\par The potential applications of the successful research outcome can be directly induced from the previously listed objectives. First of all, we would show that the general models (or the models applied in other domains) can be utilised for MSI data analysis. Further, the developed spatial smoothing methodology would improve the accuracy of the chemical processes prediction. Finally, the generalisation of the developed spatial smoothing model might suggest different approaches in tackling error correction in general; finally, the successful research outcome would suggest the applicability prospect of topic modelling in MSI data sets; reversely, it would also impact the prospect to derive machine learning generalisations from the research in MSI data sets.

@@

% Intro
% - The reason why the literature review is conducted.
% - Outline: preliminaries, notation, literature.
% - A notification the the results will be discussed in the following section.
\par The provided literature review on topic modelling is directed towards familiarising with LDA derivatives. Also, the review assesses the methodology of LDA-like models and proposes several implementation variations. The review is structured in progressing order: at the start, we look into the original LDA paper and define the terminology (it will be used throughout the proposal); then, we familiarise with two different techniques used for the inference of LDA-like models; finally, we review LDA derivatives which shows the potential to be utilised in designing the error correction technique.

@@

% General intuition on topic modelling
% - Example
\par Recall that topic modelling infers hidden patterns from the processed data sets. For example, if we were analysing a book, its chapters might address different topics or address the topics at different extent. Further, the chapters itself would have unique distributions of the topics. Note that we are assuming that there is no knowledge about the actual topics covered in the book. In order to derive these hidden topics, a probabilistic topic modelling method would analyse the structure of the words used in the book. In other words, the model would induce patterns which would refer to a particular topic. Regarding the method's performance, it would depend on the method's ability to optimise the inference of the patterns. For example, commonly used words (such as and, or, the, etc.) would be relevant in all topics. Therefore, their semantic impact in defining the topics would be negligible. For this reason, we could dismiss these to improve the time performance of the method.

% LDA intro
% - the model
% - A notification that the terminology and the notation will be useful throughout the proposal.
\par The terminology used in this proposal is similar to the terminology used by Blei \cite{blei_2012}. Figure \ref{fig:lda} below represents the original LDA model in plate notation. We will use it as a guide to familiarise with the terminology.   
\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{lda_model}
  \caption{The design of the original LDA model}
  \label{fig:lda}
\end{figure}

% High level LDA terminology
% - Notation:
% -- List LDA definitions
% -- Explain the definitions
\par To start with, we will familiarise with the higher level entities of the model. The circles represent an entity, whereas the plates indicate the number of entities. For example, there would be $N \times M$ entities of $z$ and $M$ entities of $\theta$. Further, LDA is a three-level hierarchical structure: the smallest entity (grey circle) is defined as a \textit{word}; then a collection of words (inner plate) is defined as a \textit{document}; and a collection of documents (outer plate) is a \textit{corpus}. Note that Figure \ref{fig:lda} represents $M$ documents and $N$ words per document, it follows that in the corpus there would be $M \times N$ words. Also, as suggested by Figure \ref{fig:lda}, let's denote a word by $w$; then, the sequence of words in document $d$ is denoted by 
\begin{equation*}
\mathbf{w}_d = \{w_1, w_2, \dots, w_N\},
\end{equation*}
and the sequence of documents in the corpus is denoted by
\begin{equation*}
\mathcal{\mathbf{D}} = \{\mathbf{w_1}, \mathbf{w_2}, \dots, \mathbf{w_M}\}.
\end{equation*}

% Low level LDA terminology
% - Intro to observable and hidden variables
% - Notation:
% -- List the definitions
% -- Explain the definitions
\par Now, we will introduce the lower level entities. The grey circle indicates the \textit{observable} variables, whereas the white coloured circles indicate the \textit{hidden} variables. Note that in Figure 1 the words are the only observable variables; all other entities are hidden, i.e., they describe the underlying structure of the model. Further, recall that topic modelling is applied in order to discover a mixture of topics. In other words, topic modelling could describe a corpus and its documents in terms of topic distributions. This intuition will simplify the definitions of the remaining entities in Figure 1, these are given as follows:
\begin{itemize}
\item $K$ is the number of topics;
\item $V$ is the size of the vocabulary, i.e. the number of unique words in the corpus;
\item $\alpha$ is the parameter referring to the prior distribution of the topics over documents;
\item $\beta$ is the parameter referring to the prior distribution of the vocabulary over topics;
\item $\theta$ is the latent variable referring to the topic distribution over documents;
\item $\phi$ is the latent variable referring to the vocabulary over topics
\item $w$ is the observable variable referring to the words in documents;
\item $z$ is the latent variable of topic assignments over the words.
\end{itemize}
Further, note that the listed entities can be perceived as matrices of the following dimensions: $\alpha$ is $1 \times K$; $\beta$ is $K \times V$; $\theta$ is $M \times K$; $\phi$ is $K \times V$; $w$ is $M \times N$; and $z$ is $M \times V$.
% To-do: check the dimensions

@@

% Basic LDA
% - Generative process
% - The derivation of the posterior
\par The relationship of observable and hidden variables, effectively, describes the inference of the underlying topic structure. We will go through an arbitrary example in order to give an intuition in the inference process; also, we will familiarise with the lower level notation of the previous listed variables. Figure \ref{fig:lda} illustrates that $\alpha_k$ (the prior for the $k$th topic in a document) influences $\theta_{:, k}$ (the latent variable describing the distribution of topic $k$ over all documents). It follows that $\theta_{d, k}$ (the probability of the $k$th topic occurrence in document $d$) influences $z_{d, :}$ (the topic assignment over all vocabulary terms for document $d$). Further, $z_{d, v}$ refers to the topic assignment of vocabulary term $v$. By taking $z_{d, v}$ and $\beta_{v, k}$ (the prior of topic $k$ on vocabulary term $v$), we obtain word $w_{d, v}$. In the complete perspective, the generative process of the corpus is given by the following joint probability distribution
\begin{equation}
p(\beta, \theta, z, w) = \prod_{k=1}^Kp(\beta_{:, k})\prod_{d=1}^Dp(\theta_{d, :})\bigg(\prod_{v=1}^Vp(z_{d, v} | \theta_{d, :})p(w_{d, v} | \beta, z_{d, v})\bigg).
\end{equation}
By using Bayes Theorem, we derive the computation of the conditional distribution
\begin{equation}
p(\beta, \theta, z | w) = \frac{p(\beta, \theta, z, w)}{p(w)}. 
\end{equation}
This equation expresses the computation of the posterior -- the derivation of the hidden corpus structure given observable variable $w$. The problem arises from inefficient analytical computation of evidence $p(w)$. The approximation of it will be considered during the literature review on LDA-like models. Note that the naming of the following subsections corresponds to the reviewed papers.

@@

% Assumptions 
% - Bag of words principle
% - Discrete data sets
% - Topic dimensionality
\par The original LDA model sets some assumptions for the data processed by the model. First of all, it is assumed that documents and words are \textit{exchangeable}. That is, the order in which these entities are processed does not matter. In other words, we can say the model follows the \textit{bag-of-words} principle. The next assumption states that we should use discrete data. Effectively, the basic LDA does not cover  continuous features. Another assumption is on the number of topics: it is set to be fixed throughout the complete run of inference.

@@

% Variational inference 
% - Variational parameters gamma and fi
\par As suggested by the last paragraph of the preliminaries subsection, the inference relates to the computation of evidence $p(w)$ (it is the denominator term given in Equation 2). Even though the original paper mentions the inference methods such as Markov chain Monte Carlo, the authors suggest using the method of variational inference. In order to develop an intuition over the method, the authors introduce the variational parameters $\gamma$ and $\phi$. It is also mentioned that the computation of $\gamma$ and $\phi$ is an optimisation problem. That is, we are assessing the minimum bound over the latter parameters. Further, recall that in Figure \ref{fig:lda} we explicitly express the parameter $\phi$ in describing the basic LDA model. However, the authors have chosen to show it separately -- upon describing the variational method. Their expression is given in Figure \ref{fig:var_par} below.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.4\textwidth]{variational_parameters}
  \caption{Variational parameters}
  \label{fig:var_par}
\end{figure}
Note that the authors provide pseudo code to build an intuition in tackling this optimisation problem.

% Parameter estimation
% - M-step is an update of beta
% - Newton--Raphson is used to update alpha
% - The smoothing issue: 
% -- Sparsity from large data sets
% -- The incoorperation of documents which were not used for the training
\par Additionally, the authors emphasise the estimation of the parameters $\alpha$ and $\beta$. They use a two-step EM procedure which involves the previously introduced variational parameters $\gamma$ and $\phi$. E-step is the solution of the variational parameters optimisation problem; and M-step is an update of the parameters $\alpha$ and $\beta$. Note that the authors provide an analytical approach to estimate $\beta$ which is proportional to the changes of $\phi$; and for $\alpha$, they provide an approach utilising the Newton--Raphson method. Apart from that, the authors introduce the an issue of \textit{sparsity} in a corpus structure. A sparse corpus is expected upon using a large vocabulary or a large amount of documents. This issue is addressed by introducing smoothing: an additional parameter $\eta$ is utilised to smooth the estimation of $\beta$.

% Experiments
% - Document modelling: issue with pLSI and mixture unigrams~ 
% - Document classification: 
% -- Dimensionality reduction
% -- Comparison with SVM
\par The conducted experiments display an improvement in performance over the predecessor models and suggest the application domains of LDA-like models. The first experiment addresses document modelling. To be more specific, LDA has displayed balanced hidden topic proportions in the set of tested documents. That is, the tested documents has replicated the topic proportions of the training documents. The second experiment is directed towards the problem of document classification. The authors have addressed the \textit{dimensionality reduction} in order to surpass the performance of the support vector machine (SVM) models. To be more specific, the documents with reduced dimensionality would possess a lower amount of features. The results of the experiment indicate that the basic LDA model has successfully reduced the number of features. Also, it has displayed an improvement in accuracy compared to the SVM-like models.

% Conclusion
% - Summary of the results established by the paper
\par The review on the original LDA model has set up a basis required to understand the model's derivatives. That is, the introduced methodology will be useful in considering the following improvements: (1) the  relaxation of the exchangeability assumption; (2) the adaptation of time-series suggesting the word-topic correspondence over time; (3) the discussion over alternative inference methods. Apart from that, the experiments introduced in this section have reflected the LDA performance during the time of the initial model release. The current state-of-the-hard topic modelling techniques overcome the shown results. Nevertheless, the carried experiments suggest the application fields of LDA-like models.
% To-do: rework the paragraph name to reflect its purpose: remove the discussion (you might think it's done, but it's not)

@@

% Intro
% - Sampling as a method of inference
\par The research paper titled as `Finding Scientific Topics' \cite{griffiths} suggests an alternative approach to the inference which is suitable for LDA-like models. To be more specific, the article provides a readable introduction on the application of the Markov chain Monte Carlo (MCMC) algorithm. That is, the MCMC algorithm is applied as a sampling-based method to infer an underlying topic structure. Further, the authors discuss the method's application settings for the machine learning problems in vision. Also, the authors conduct an experiment on inferring the structure of topics in the abstracts of papers published in `Proceedings of the National Academy of Sciences' (PNAS). Finally, the paper is concluded by a discussion on assessing the sampling-based method usability. To be more specific, the performance of the sampled-based method is compared to the performance of variational inference methods.

>>>>

% other potentially useful packages
\usepackage{amssymb,amsmath}
\usepackage{textcomp}
%\usepackage{url}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Probabilistic Topic Modelling for Spatial Smoothing in Mass Spectrometry Imaging}
\author{Arijus Pleska}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\par In this project, the application domain of MSI is set to be \textit{metabolomics} -- it is the scientific study of chemical processes which involve metabolites (small molecules) and metabolomes (sets of small molecules). In the research settings of metabolomics, we utilise \text{mass spectrometry} (MS) to obtain the distribution of chemical entities within a sample. By utilising MSI, we can visualise the distributions of the ions (or ion sets) of different masses. For instance, Figure \ref{fig:b_and_e} below represents a captured sample as an image, where each picture's pixel is a small region within the sample. 
\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{b_and_e}
  \caption{MSI application on a sample with the letters b and e.}
  \label{fig:b_and_e}
\end{figure}
Note that each pixel contains a distribution of chemical entities as well. In Figure \ref{fig:b_and_e} we have chosen the range of mass to reflect the ink used to inscribe the letters. That is, for each pixel, we have summed the intensities of the chemical entities meeting the mass criteria. Then, each pixel has been assigned an intensity value. Therefore, the sample could be expressed as the latter image. An open problem in metabolomics is to find the sets of chemical entities  inducing valid visual patterns. Ultimately, the discovered biological patterns could be used to suggest and measure the likelihood of a particular chemical process occurrence. For a rigorous discussion on machine learning applications in metabolomics, the reader can consult the recent survey conducted by Alonso et al. \cite{alonso-et-al}.

@@

% Unsupervisder learning
% - Topic moels
% - Probabilistic models
% - LDA:
% -- General
% -- Metabolomics
\par Since this proposal is focused on utilising spatial smoothing to reduce the impact of noise in MSI data sets, we will familiarise with the machine learning techniques and problems tackled in the MSI domain. The objective to discover novel patterns can be treated as a problem of \textit{unsupervised} machine learning. That is, we have no initial settings in how the patterns are expected to look like. However, upon appropriate optimisation, the unsupervised machine learning techniques are capable to learn the patterns on their own. To be more specific, we will utilise the techniques of \textit{topic modelling} (in this case, a pattern is referred by a topic). Also, we are focusing on the \textit{probabilistic} (statistical) models, more specifically, \textit{latent Dirichlet allocation} (LDA). First of all, the successor LDA-like models are treated as a state-of-the-art method in tackling semantic analysis type problems. For more detailed review on LDA applications, the reader can consult the survey conducted by Blei \cite{blei_2012}. Also, Hooft et al. \cite{hooft} has recently addressed the applicability of LDA in MSI data sets. This proposal is directly influenced by the latter study: we will utilise the LDA-like models to reduce the impact of noise in the MSI data sets.

@@

% Objectives
% - Research in the state-of-the-art methodology.
% - Reducing the impact of noise in metabolomics.
% - Providing a general-purpose machine learning model.
\par The primary goal of this research is to expand the consensus on whether topic modelling is a valid approach to apply spatial smoothing in MSI data sets. To be more specific, we set our objectives to the following:
\begin{enumerate}
    \item Optimise the general topic modelling methods for MSI data sets;
    \item Successfully apply spatial smoothing in MSI data sets;
    \item Give a basis for the development of a general topic modelling method for error correction.
\end{enumerate}
Briefly, for the first objective, we will conduct a survey on LDA-like topic modelling methodology. This survey will distinguish the potential models for the applications in MSI. The second objective focuses on optimising these potential models for the particular task of spatial smoothing in MSI data sets. Finally, the third objective can be treated as a general contribution to the field of machine learning -- the discovered spatial smoothing techniques might be applicable to other domains.

% Hypothesis
% - Topic modelling can improve the quality of metabolomics data sets.
% - The negation of the hypothesis would contribute in terms of reassessing the complexity of the problem.
\par The main project hypothesis -- the LDA-like models can detect and correct the degeneracies of MSI data sets -- can be interpreted as a spatial smoothing application in MSI data sets. Both cases of proving and disproving this hypothesis would contribute to the research in MSI. The successful outcome would lead to the potential applications discussed in the following paragraph, whereas the unsuccessful outcome would set up additional requirements to tackle the issue of noisiness in MSI data sets.

% Results: tools to annotate the metabolites > basis for generative error correction 
% - Expanding the application of existing machine learning models.
% - Advancing the current state-of-the-art in metabolomics.
% - Contributing to machine learning in general.
\par The potential applications of the successful research outcome can be directly induced from the previously listed objectives. First of all, we would show that the general models (or the models applied in other domains) can be utilised in MSI data analysis. Further, the developed spatial smoothing methodology would improve the accuracy of the chemical processes prediction. Finally, the generalisation of the developed spatial smoothing model might suggest different approaches in tackling error correction in general; finally, the successful research outcome would suggest the applicability prospect of topic modelling in MSI data sets; reversely, it would also impact the prospect to derive machine learning generalisations from the research in MSI data sets.

@@

% Intro
% - The reason why the literature review is conducted.
% - Outline: preliminaries, notation, literature.
% - A notification the the results will be discussed in the following section.
\par The provided literature review on topic modelling is directed towards familiarising with LDA derivatives. Also, the review assesses the methodology of LDA-like models and proposes several implementation variations. The review is structured in progressing order: at the start, we look into the original LDA paper and define the terminology (it will be used throughout the proposal); then, we familiarise with two different techniques used for the inference of LDA-like models; finally, we review LDA derivatives which show the potential to be utilised as a technique of error correction.

@@

% General intuition on topic modelling
% - Example
\par Recall that topic modelling infers hidden patterns from the processed data sets. For example, if we were analysing a book, its chapters might address different topics or address the topics at different extent. Further, the chapters itself would have unique distributions of the topics. Note that we are assuming that there is no knowledge about the actual topics covered in the book. In order to derive those hidden topics, a probabilistic topic modelling method would analyse the structure of the words used in the book. In other words, the model would induce patterns which would refer to a particular topic. The method's performance would depend on its ability to optimise the inference of the patterns. For example, commonly used words (and, or, the, etc.) would be relevant in all topics. Therefore, their semantic impact in defining the topics would be negligible. For this reason, we could dismiss those words to improve the time performance of the method.

% LDA intro
% - the model
% - A notification that the terminology and the notation will be useful throughout the proposal.
\par The terminology used in this proposal is similar to the terminology used by Blei \cite{blei_2012}. Figure \ref{fig:lda} below represents the original LDA model in plate notation. We will use it as a guide to familiarise with the terminology.   
\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{lda_model}
  \caption{The design of the original LDA model.}
  \label{fig:lda}
\end{figure}

% High level LDA terminology
% - Notation:
% -- List LDA definitions
% -- Explain the definitions
\par To start with, we will familiarise with the higher level entities of the model. The circles represent particular entities, whereas the plates indicate the number of entities. For example, there would be $N \times M$ entities of $z$ and $M$ entities of $\theta$. Further, LDA is a three-level hierarchical structure: the smallest entity (grey circle) is referred by a \textit{word}; then a collection of words (inner plate) is referred by a \textit{document}; and a collection of documents (outer plate) is referred by a \textit{corpus}. Note that Figure \ref{fig:lda} represents $M$ documents and $N$ words per document, it follows that in the corpus there would be $M \times N$ words. Also, as suggested by Figure \ref{fig:lda}, let's denote a word by $w$; then, the sequence of words in document $d$ would be denoted by 
\begin{equation*}
\mathbf{w}_d = \{w_1, w_2, \dots, w_N\},
\end{equation*}
and the sequence of documents in the corpus would be denoted by
\begin{equation*}
\mathcal{\mathbf{D}} = \{\mathbf{w_1}, \mathbf{w_2}, \dots, \mathbf{w_M}\}.
\end{equation*}

% Low level LDA terminology
% - Intro to observable and hidden variables
% - Notation:
% -- List the definitions
% -- Explain the definitions
\par Now, we will introduce the lower level entities. The grey circle indicates the \textit{observable} variables, whereas the white coloured circles indicate the \textit{hidden} variables. Note that in Figure 1 the words are the only observable variables; all other entities are hidden, i.e., they describe the underlying structure of the model. Further, recall that topic modelling is applied in order to discover a mixture of topics. In other words, topic modelling could describe a corpus and its documents in terms of topic distributions. This intuition will simplify the definitions of the remaining entities in Figure 1, these are given as follows:
\begin{itemize}
\item $K$ is the number of topics;
\item $V$ is the size of the vocabulary, i.e. the number of unique words in the corpus;
\item $\alpha$ is the parameter referring to the prior distribution of the topics over documents;
\item $\beta$ is the parameter referring to the prior distribution of the vocabulary over topics;
\item $\theta$ is the latent variable referring to the topic distribution over documents;
\item $\phi$ is the latent variable referring to the vocabulary distribution over topics
\item $w$ is the observable variable referring to the words in documents;
\item $z$ is the latent variable of topic assignments over the words.
\end{itemize}
Further, note that the listed entities can be perceived as matrices of the following dimensions: $\alpha$ is $1 \times K$; $\beta$ is $K \times V$; $\theta$ is $M \times K$; $\phi$ is $K \times V$; $w$ is $M \times N$; and $z$ is $M \times N$.

@@

% Basic LDA
% - Generative process
% - The derivation of the posterior
\par The relationship of observable and hidden variables, effectively, describes the inference of the underlying topic structure. We will go through an arbitrary example in order to give an intuition in the inference process; also, we will familiarise with the lower level notation of the previous listed variables. Figure \ref{fig:lda} illustrates that $\alpha_k$ (the prior for the $k$th topic in a document) influences $\theta_{:, k}$ (the latent variable describing the distribution of topic $k$ over all documents). It follows that $\theta_{d, k}$ (the probability of the $k$th topic occurrence in document $d$) influences $z_{d, :}$ (the topic assignments of of the words in document $d$). More specifically, $z_{d, v}$ refers to the topic assignment of vocabulary term $v$. By taking $z_{d, v}$ and $\beta_{k, v}$ (the prior of topic $k$ on vocabulary term $v$), we obtain word $w_{d, v}$. In the complete perspective, the generative process of the corpus is given by the following joint probability distribution
\begin{equation}
p(\beta, \theta, z, w) = \prod_{k=1}^Kp(\beta_{:, k})\prod_{d=1}^Dp(\theta_{d, :})\bigg(\prod_{v=1}^Vp(z_{d, v} | \theta_{d, :})p(w_{d, v} | \beta, z_{d, v})\bigg).
\end{equation}
By using Bayes Theorem, we derive the computation of the conditional distribution
\begin{equation}
p(\beta, \theta, z | w) = \frac{p(\beta, \theta, z, w)}{p(w)}. 
\end{equation}
This equation expresses the computation of the posterior -- the derivation of the hidden corpus structure given observable variable $w$. The problem arises from inefficient analytical computation of evidence $p(w)$. The approximation of $p(w)$ will be considered during the literature review on LDA-like models. Note that the naming of the following subsections corresponds to the reviewed papers.

@@

% Assumptions 
% - Bag of words principle
% - Discrete data sets
% - Topic dimensionality
\par The original LDA model sets some assumptions for the data processed by the model. First of all, it is assumed that documents and words are \textit{exchangeable}. That is, the order in which these entities are processed does not matter. In other words, the model follows the \textit{bag-of-words} principle. The next assumption states that we should use discrete data. Therefore, note that the basic LDA model does not cover continuous features. Another assumption is on the number of topics: it is set to be fixed throughout the complete run of inference.

@@

% Variational inference 
% - Variational parameters gamma and fi
\par As suggested by the last paragraph of the preliminaries subsection, the inference relates to the computation of evidence $p(w)$ (it is the denominator term given in Equation 2). Even though the original paper mentions the inference methods such as Markov chain Monte Carlo, the authors suggest using the method of variational inference. In order to develop an intuition over the method, the authors introduce the variational parameters $\gamma$ and $\phi$. It is also mentioned that the computation of $\gamma$ and $\phi$ is an optimisation problem. That is, we are assessing the minimum bound over the latter parameters. Further, recall that in Figure \ref{fig:lda} we explicitly express the parameter $\phi$ in describing the basic LDA model. However, the authors have chosen to show it separately -- upon describing the variational method. Their expression is given in Figure \ref{fig:var_par} below.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.4\textwidth]{variational_parameters}
  \caption{Variational parameters.}
  \label{fig:var_par}
\end{figure}
Note that the authors provide pseudo code to develop an intuition in tackling this optimisation problem.

% Parameter estimation
% - M-step is an update of beta
% - Newton--Raphson is used to update alpha
% - The smoothing issue: 
% -- Sparsity from large data sets
% -- The incoorperation of documents which were not used for the training
\par Additionally, the authors emphasise the estimation of the parameters $\alpha$ and $\beta$. They use a two-step EM procedure which involves the previously introduced variational parameters $\gamma$ and $\phi$. E-step is the solution of the variational parameters optimisation problem; and M-step is an update of the parameters $\alpha$ and $\beta$. Note that the authors provide an analytical approach to estimate $\beta$ which is proportional to the changes of $\phi$; and for $\alpha$, they provide an approach utilising the Newton--Raphson method. Apart from that, the authors introduce an issue of \textit{sparsity} in the corpus structure. A sparse corpus is expected upon using a large vocabulary or a large amount of documents. This issue is addressed by introducing smoothing: an additional parameter $\eta$ is utilised to smooth the estimation of $\beta$.

% Experiments
% - Document modelling: issue with pLSI and mixture unigrams~ 
% - Document classification: 
% -- Dimensionality reduction
% -- Comparison with SVM
\par The conducted experiments display an improvement in performance over the predecessor models and suggest the application domains of LDA-like models. The first experiment addresses document modelling. To be more specific, LDA has displayed balanced hidden topic proportions in the set of tested documents. That is, the tested documents have replicated the topic proportions of the training documents. The second experiment is directed towards the problem of document classification. The authors have addressed the \textit{dimensionality reduction} in order to surpass the performance of the support vector machine (SVM) models. To be more specific, the documents with reduced dimensionality would possess lower amount of features. The results of the experiment indicate that the basic LDA model has successfully reduced the number of features. Also, it has displayed an improvement in accuracy compared to the SVM-like models.

% Conclusion
% - Summary of the results established by the paper
\par The review on the original LDA model has set up a basis required to understand the model's derivatives. That is, the introduced methodology will be useful in considering the following improvements: (1) the  relaxation of the exchangeability assumption; (2) the adaptation of time-series suggesting the word-topic correspondence over time; (3) the discussion over alternative inference methods. Apart from that, the experiments introduced in this section have reflected the LDA performance during the time of the initial model release. Current state-of-the-art techniques overcome the shown results. Nevertheless, the carried experiments suggest the application fields of LDA-like models.
% To-do: rework the paragraph name to reflect its purpose: remove the discussion (you might think it's done, but it's not)

@@

% Intro
% - Sampling as a method of inference
\par The research paper titled as `Finding Scientific Topics' \cite{griffiths} suggests an alternative approach to the inference which is suitable for LDA-like models. To be more specific, the article provides a readable introduction on the application of the Markov chain Monte Carlo (MCMC) algorithm. That is, the MCMC algorithm is applied as a sampling-based method to infer an underlying topic mixture. Further, the authors discuss the method's application settings for the machine learning problems in vision. Also, the authors conduct an experiment on the topic inference in the abstracts of the research papers published in `Proceedings of the National Academy of Sciences' (PNAS). Finally, the paper is concluded by assessing  the sampling-based method's usability. To be more specific, the performance of the sampled-based method is compared to the performance of variational inference methods.

% Gibbs Sampling
% - Recall the meaning of phi and theta.
% - Introduce the inference method
\par The MCMC-based inference method is often referred by \textit{Gibbs sampling}. The authors claim that the method provides a first-order approximation for the inference of the hidden structure of the data. That is, the performance is sufficient to establish quantitative reasoning in assessing the correlating sections and semantic structure of the data. Recall that by $\theta$ we denote the topic distribution over the documents and by $\phi$ we denote the vocabulary terms distribution over the topics. In Gibbs sampling, the latter variables are obtained by examining the posterior distribution, whereas topic assignments $z$ are sampled sequentially. That is, we sample topic assignment $z$ and update the distribution depending on the obtained value (this will impact the next sample of $z$); we repeat this process until the topic distribution converges. To be more specific, each drawn $z_i$ ($i$ denotes the iteration) is recorded as a count. That is, we keep a track of topic assignments for particular words and documents. The value of drawn topic $z_i$ is proportional to the distribution given below:
\begin{equation}
P(z_i = j | z_{-i}, w) \propto \dfrac{n_{-i, j}^{(w_i)} + \beta}{n_{-i, j}^{(\cdot)} + V\beta}\dfrac{n_{-i, j}^{(d_i)} + \alpha}{n_{-i, \cdot}^{(d_i)} + K\alpha}.
\end{equation}
For a brief familiarisation with the notation used in the expression, note the following terminology: $n_j^{d}$ and $n_j^{w}$ are the counts of topic $j$ assignments in document $d$ and word $w$, respectively; recall that $V$ is the number of words in the vocabulary and $K$ is the number of topics. Also, note that the counts do not include the last assignment of the topic, this is indicated by $-i$. Ultimately, the left side of the expression is $\phi$ and the right side is $\theta$. That is, 
\begin{align}
\phi_j &= \dfrac{n_{-i, j}^{(w_i)} + \beta}{n_{-i, j}^{(\cdot)} + W\beta},\\
\theta_j &= \dfrac{n_{-i, j}^{(d_i)} + \alpha}{n_{-i, \cdot}^{(d_i)} + K\alpha}.
\end{align}
This process of the topic assignment is executed for every word in the document and for all documents in the corpus. The procedure is repeated until convergence or, more pragmatically, until the distribution displays the assigned topic fluctuations below the set threshold. In the Gibbs sampling procedure, there is no need to explicitly assign $\theta$ and $\phi$ values. If needed, these values are obtained directly from the topic distribution, as shown by the two previous equations.

% Auxilliary experiments
% - Vision
% - Time series
\par The authors have shown an application of Gibbs sampling in a problem of vision. In their experiment settings, it is assumed that a collection of images represents a corpus; an image represents a document; and a pixel of an image is a word. The experiment has shown that the model is able to infer the underlying basis (topics) from which the images could be constructed. The performance of the model was compared to the variational inference based LDA models. Note that the performance was assessed in terms of \textit{perplexity} -- the uncertainty in assigning a word to a topic. It has also been shown that Gibbs sampling displays superior performance in small data sets, whereas the performance in larger data sets tends to fall off. Furthermore, the authors have assessed the performance upon varying the number of topics. The results of this experiment indicate a trade-off in deriving high-level and low-level patterns. That is, the low number of topics might result in mashing significant patterns, whereas the high number of topics is likely to infer irrelevant patterns.

% The main experiment
% - PNAS
\par The main experiment of the research paper is conducted on the PNAS data set containing abstracts of the articles published from 1991 to 2001. The authors have noticed that the topic distribution over the vocabulary terms change over time. However, the model had not yet been optimised to take this notion of time-series into account upon the inference. Nevertheless, the authors have emphasised the necessity to assess the dynamic topic modelling in the future.
